---
title: "Mental Health in Newfoundland and Labrador"
author: "John W"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

# Introduction

This notebook makes use of several datasets published by the Government of Canada on mental health: 

- [here](https://open.canada.ca/data/en/dataset/480a161e-56c6-4a86-869c-e38381b329a5) 
- [here](https://open.canada.ca/data/en/dataset/a9863f45-6a1a-4277-ae0f-2d9cb61d413a)

The purpose of this notebook is to provide a general survey of mental health in Newfoundland and Labrador, Canada.

## Analysis

## How do the people of Newfoundland and Labrador rate their own mental health?

```{r}
  library(tidyverse)
  library(viridis)
  library(janitor)
  
  data <- read.csv("13100085.csv")
  data <- clean_names(data)
  
  df <- data %>%
    filter(
      age_group != "Total, 12 years and over",
      sex != "Both sexes",
      geo == "Newfoundland and Labrador",
      characteristics == "Number of persons",
      !grepl("Total population", self_rated_mental_health, ignore.case = TRUE),
      self_rated_mental_health != "Very good or excellent self-rated mental health"
    )
  
  palette <- viridis(length(unique(df$self_rated_mental_health)))

  male <- df %>% filter(sex == "Males")
  
  ggplot(male, aes(x = age_group, y = value, fill = self_rated_mental_health)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = palette) +
    labs(
      title = "Self-Rated Mental Health by Age Group (Males)",
      subtitle="Newfoundland and Labrador",
      x = "Age Group",
      y = "Number of Persons",
      fill = "Mental Health"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
  
  female <- df %>% filter(sex == "Females")
  
  ggplot(female, aes(x = age_group, y = value, fill = self_rated_mental_health)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = palette) +
    labs(
      title = "Self-Rated Mental Health by Age Group (Females)",
      subtitle = "Newfoundland and Labrador",
      x = "Age Group",
      y = "Number of Persons",
      fill = "Mental Health"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
```

Males in Newfoundland and Labrador generally have a positive mental health outlook. The vast majority of respondents across all age groups report their mental health as being between good and excellent, with few (if any) individuals rating their mental health as fair or poor. The 65 and over age group notably has a number of non-respondents. Generally, however, the outlook of males in the province is a positive one. The smallest number of respondents came from the 12-19 age group, whereas the largest number of respondents came from the 45-64 age group, indicating a strong concentration of responses among middle-aged respondents.

Similarly, females seem to have a generally positive mental health outlook. The vast majority of females in the province report their mental health as being good, very good or excellent. However, a disparity exists in the 45-64 and 65 and over age groups, in which there are a number of respondents who report their mental health as being fair or poor. This could potentially indicate a decline in mental health outlook with age. As in the case of males, the smallest number of respondents came from the 12-19 age group, with the largest number of respondents from the 45-64 age group.

## What is the perceived need for mental health services in Newfoundland and Labrador?

```{r}
  library(tidyverse)
  library(janitor)
  library(viridis)

  options(scipen=999)
  
  indicators_df <- read.csv("13100465.csv")

  indicators_df <- indicators_df |>
    janitor::clean_names() |>
    filter(str_detect(indicators, regex("Perceived need for mental health care", ignore_case = TRUE))) |>
    mutate(indicators = str_remove(indicators, "Perceived need for mental health care, ")) |>
    filter(
      geo=="Newfoundland and Labrador",
      age_group!="Total, 15 years and over",
      gender!="Total, gender of person",
      characteristics=="Number of persons",
      uom=="Number",
      ref_date=="2012",
      age_group!="25 to 64 years"
    )
  
  indicators_df$gender <- str_remove(indicators_df$gender, "\\+$")
  
  palette <- viridis(length(unique(indicators_df$indicators)))
  
  male <- indicators_df |>
    filter(gender=="Men") |>
    ggplot(aes(x=age_group,y=value,fill=indicators)) +
    geom_col() +
    scale_fill_manual(values = palette) +
    labs(
      title="Need for Mental Health Services (Men)",
      subtitle="Newfoundland and Labrador (2012)",
      x="Year",
      y="Number of Persons",
      fill="Need"
    )
  
  male
  
  female <- indicators_df |>
    filter(gender=="Women") |>
    ggplot(aes(x=age_group,y=value,fill=indicators)) +
    geom_col() +
    scale_fill_manual(values = palette) +
    labs(
      title="Need for Mental Health Services (Women)",
      subtitle="Newfoundland and Labrador (2012)",
      x="Year",
      y="Number of Persons",
      fill="Need"
    )
  
  female
```

The overall mental health service need appears low across all age groups. "No need" dominates in every category, making up the vast majority of each bar. The highest number of respondents came from the 45-64 age group. The smallest number of respondents came from the 15-24 age group. Each of the 15-24, 25-44 and 45-64 age groups recorded a small portion of respondents who stated that their mental health needs were met. There
were evidently no reports of needs being partially met or not met. The majority of men in the province report adequate mental health care or no perceived need for care.

The situation is slightly different for women in Newfoundland and Labrador. The vast majority of women across all age groups reported no need for mental health care services. The exception is the 45-64 age group, which reported both a portion of respondents who stated all their needs were met, as well as a noticeable portion of respondents who stated their needs were only partially met. This could indicate a need for increased mental health care services for women, particularly in older age groups. The largest number of respondents came from the 45-64 age group; the smallest number of respondents came from the 15-24 age group.

# Conclusion

Overall, mental health outlook is positive for both males and females in Newfoundland and Labrador. Most individuals report good to excellent mental health, with very few reporting poor outcomes. Males in the province have a consistently strong outlook across all age groups. There is minimal evidence of unmet mental health needs and the majority report no need for services or adequate care. In the case of females outlook is generally positive, but slightly more variable with age. Older age groups show higher reports of fair/poor mental health. There is some evidence of partially met needs, particularly for middle-aged women. The need for services appears to be low across the overall population, as "no need" is the dominant response for both genders when asked about their personal need for mental health care. However, women (especially 45-64) show signs of greater unmet or partially met needs. The 45-64 age group has the highest number of respondents and is the most informative; younger groups are comparatively underrepresented. While the general outlook is positive, there are emerging gaps in mental health support for older women, suggesting a need for more targeted services in this group.